---
title: "Duckeed 2019 - macroalgae"
author: "Antonino Malacrinò, edited LB"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
code_download: yes
---
```{r, warning=FALSE, message=FALSE, echo=FALSE}
library("phyloseq")
library("ggplot2")
library("scales")
library("grid")
library("DESeq2")
library("vegan")
library("ape")
library("plyr") 
library("ggpmisc")
library("dplyr")
library("broom")
library("picante")
library("Rmisc")
library("emmeans")
library("car")
library("lme4")
library("tibble")
library("data.table")
library("limma")
library("microbiome")
library("ggvenn")
library("patchwork")
library("RColorBrewer")
library("RVAideMemoire")
library("data.table")
library("decontam")
library("ggrepel")

NTHREADS = 8

create_dt <- function(x){
  DT::datatable(x,
                extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                               buttons = c('copy', 'csv', 'excel'),
                               lengthMenu = list(c(10,25,50,-1),
                                                 c(10,25,50,"All"))))
}
```


```{r, warning=FALSE, message=FALSE, echo=FALSE}
load(file = 'data/asv_its.rds')
ps.its <- ps
load(file = 'data/asv_18s.rds')
ps.18s <- ps
#metadata <- read.table("data/metadata_its.txt", sep = "\t", header = T, row.names = 1)
#sample_data(ps.its) <- metadata

remove.cont <- function(GM){
  sample_data(GM)$is.neg <- sample_data(GM)$treatment == "neg"
  contamdf.prev <- isContaminant(GM, method="prevalence", neg="is.neg", threshold = 0.05)
  cont.remove <- subset(contamdf.prev, contaminant == "TRUE")
  cont.remove <- row.names(cont.remove)
  allTaxa = taxa_names(GM)
  allTaxa <- allTaxa[!(allTaxa %in% cont.remove)]
  GM <-  prune_taxa(allTaxa, GM)
  GM <- subset_samples(GM, treatment != "neg")
  GM <- subset_samples(GM, treatment != "posContr")
  return(GM)
}

ps.its <- remove.cont(ps.its)
ps.its <- subset_taxa(ps.its, Order !="Chloroplast")
ps.its <- subset_taxa(ps.its, Family !="Mitochondria")
ps.its <- filter_taxa(ps.its, function (x) {sum(x > 0) > 1}, prune=TRUE)

ps.18s <- remove.cont(ps.18s)
ps.18s <- subset_taxa(ps.18s, Class !="Streptophyta")
ps.18s <- subset_taxa(ps.18s, Phylum !="Opisthokonta")
ps.18s <- subset_taxa(ps.18s, Phylum =="Archaeplastida")
ps.18s <- filter_taxa(ps.18s, function (x) {sum(x > 0) > 1}, prune=TRUE)
#x <- as.data.frame(tax_table(ps.18s))
```

# General stats - ITS {.tabset}
## General
```{r, warning=FALSE, message=FALSE, echo=FALSE}
ps.its
```

## Seq depth
```{r, warning=FALSE, message=FALSE, echo=FALSE}
sdt = data.table(as(sample_data(ps.its), "data.frame"),
                 TotalReads = sample_sums(ps.its), keep.rownames = TRUE)
setnames(sdt, "rn", "SampleID")
sdt
```

## Rarefaction curves
```{r, warning=FALSE, message=FALSE, echo=FALSE}
#rarecurve((otu_table(ps.its)), step=500, cex=0.5)
```


# General stats - 18S {.tabset}
## General
```{r, warning=FALSE, message=FALSE, echo=FALSE}
ps.18s
```

## Seq depth
```{r, warning=FALSE, message=FALSE, echo=FALSE}
sdt = data.table(as(sample_data(ps.18s), "data.frame"),
                 TotalReads = sample_sums(ps.18s), keep.rownames = TRUE)
setnames(sdt, "rn", "SampleID")
sdt
```

## Rarefaction curves
```{r, warning=FALSE, message=FALSE, echo=FALSE}
#rarecurve((otu_table(ps.18s)), step=500, cex=0.5)
```

# Multivariate ITS {.tabset}
## PERMANOVA
```{r, warning=FALSE, message=FALSE, echo=FALSE}
sampledf <- data.frame(sample_data(ps.its))
dist.mat <- phyloseq::distance(ps.its, method = "bray")
perm <- how(nperm = 999)
setBlocks(perm) <- with(sampledf, pond)
set.seed(100)
pmv <- adonis2(dist.mat ~ treatment, data = sampledf, permutations = perm)
pmv
```

## NMDS
```{r, warning=FALSE, message=FALSE, echo=FALSE}
cap_ord <- ordinate(physeq = ps.its, method = "NMDS", distance = dist.mat, formula = ~ 1)
cap_plot <- plot_ordination(physeq = ps.its, ordination = cap_ord, axes = c(1,2)) +
  theme_bw(base_size = 20) +
  stat_ellipse(mapping = aes(color = treatment),
               alpha = 0.4,
               type = "norm",
               show.legend=F) +
  geom_point(mapping = aes(color = treatment), size = 5) +
  theme(legend.title= element_blank(), 
        legend.background = element_rect(color = NA),
        legend.text = element_text(size = 12),
        axis.text.x = element_text(color="black"),
        axis.text.y = element_text(color="black"),
        panel.grid = element_blank()) +
  scale_color_manual(name = "Legend", values=c("#1b9e77", "#d95f02"), labels = c("Control", "Herbivory"), breaks = c("ctr", "herb"))
ggsave(cap_plot, filename = "figures/nmds_ITS.pdf", dpi = 600,  width = 8, height = 5, units = "in")
cap_plot
```

# Multivariate 18S {.tabset}
## PERMANOVA
```{r, warning=FALSE, message=FALSE, echo=FALSE}
sampledf <- data.frame(sample_data(ps.18s))
dist.mat <- phyloseq::distance(ps.18s, method = "bray")
perm <- how(nperm = 999)
setBlocks(perm) <- with(sampledf, pond)
set.seed(100)
pmv <- adonis2(dist.mat ~ treatment, data = sampledf, permutations = perm)
pmv
```

## NMDS
```{r, warning=FALSE, message=FALSE, echo=FALSE}
cap_ord <- ordinate(physeq = ps.18s, method = "NMDS", distance = dist.mat, formula = ~ 1)
cap_plot <- plot_ordination(physeq = ps.18s, ordination = cap_ord, axes = c(1,2)) +
  theme_bw(base_size = 20) +
  stat_ellipse(mapping = aes(color = treatment),
               alpha = 0.4,
               type = "norm",
               show.legend=F) +
  geom_point(mapping = aes(color = treatment), size = 5) +
  theme(legend.title= element_blank(), 
        legend.background = element_rect(color = NA),
        legend.text = element_text(size = 12),
        axis.text.x = element_text(color="black"),
        axis.text.y = element_text(color="black"),
        panel.grid = element_blank())+
  scale_color_manual(name = "Legend", values=c("#1b9e77", "#d95f02"), labels = c("Control", "Herbivory"), breaks = c("ctr", "herb"))
ggsave(cap_plot, filename = "figures/nmds_18s.pdf", dpi = 600,  width = 8, height = 5, units = "in")
cap_plot
```

# Diversity ITS {.tabset}
## model Shannon
```{r, warning=FALSE, message=FALSE, echo=FALSE}
div <- microbiome::alpha(ps.its, index = "all")
otus <- as.data.frame((otu_table(ps.its)))
#tree <- phy_tree(ps.its)
#div.pd <- pd(otus, tree, include.root = FALSE)
div <- cbind(sample_data(ps.its), div)
#div <- cbind(div, div.pd)

model <- lmer(diversity_shannon ~ treatment * (1|pond), data = div)
Anova(model)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
div_plot <- ggplot(div, aes(x = treatment, y = diversity_shannon, fill = treatment)) +
  theme_bw(base_size = 14) +
  geom_boxplot(outlier.colour="black", notch=F, outlier.shape=NA) +
  labs(y = "Shannon diversity") +
  theme(axis.title.x=element_blank(),
        axis.text.x = element_text(color="black"),
        axis.text.y = element_text(color="black"),
        axis.title.y = element_text(color="black"),
        panel.grid = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank()) +
    theme(legend.title=element_blank(),
        legend.background = element_rect(fill="transparent"),
        legend.key = element_rect(fill="transparent"),
        legend.text = element_text(size = 10),
        legend.position = "none") +
  scale_fill_manual(name = "Legend", values=c("#1b9e77", "#d95f02"), labels = c("Control", "Herbivory"), breaks = c("ctr", "herb"))
ggsave(div_plot, filename = "figures/div_ITS_shannon.pdf", dpi = 600,  width = 8, height = 5, units = "in")
div_plot
```

# Diversity 18S {.tabset}
## model Shannon
```{r, warning=FALSE, message=FALSE, echo=FALSE}
div <- microbiome::alpha(ps.18s, index = "all")
otus <- as.data.frame((otu_table(ps.18s)))
#tree <- phy_tree(ps.18s)
#div.pd <- pd(otus, tree, include.root = FALSE)
div <- cbind(sample_data(ps.18s), div)
#div <- cbind(div, div.pd)

model <- lmer(diversity_shannon ~ treatment * (1|pond), data = div)
Anova(model)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
div_plot <- ggplot(div, aes(x = treatment, y = diversity_shannon, fill = treatment)) +
  theme_bw(base_size = 14) +
  geom_boxplot(outlier.colour="black", notch=F, outlier.shape=NA) +
  labs(y = "Shannon diversity") +
  theme(axis.title.x=element_blank(),
        axis.text.x = element_text(color="black"),
        axis.text.y = element_text(color="black"),
        axis.title.y = element_text(color="black"),
        panel.grid = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank()) +
    theme(legend.title=element_blank(),
        legend.background = element_rect(fill="transparent"),
        legend.key = element_rect(fill="transparent"),
        legend.text = element_text(size = 10),
        legend.position = "none")+
  scale_fill_manual(name = "Legend", values=c("#1b9e77", "#d95f02"), labels = c("Control", "Herbivory"), breaks = c("ctr", "herb"))
ggsave(div_plot, filename = "figures/div_18s_shannon.pdf", dpi = 600,  width = 8, height = 5, units = "in")
div_plot
```


# Taxonomical composition {.tabset}
## ITS
```{r, warning=FALSE, message=FALSE, echo=FALSE}
diagdds = phyloseq_to_deseq2(ps.its, ~ 1)
ts = counts(diagdds)
geoMeans = apply(ts, 1, function(row) if (all(row == 0)) 0 else exp(mean(log(row[row != 0]))))
diagdds = estimateSizeFactors(diagdds, geoMeans=geoMeans)
diagdds = estimateDispersions(diagdds)
diagvst = getVarianceStabilizedData(diagdds)
diagdds.c <- removeBatchEffect(diagvst)
diagdds.c[diagdds.c<0] <- 0
ps.itsn <- ps.its
otu_table(ps.itsn) <- otu_table(diagdds.c, taxa_are_rows = TRUE)

glom <- microbiome::aggregate_taxa(ps.itsn, "Genus")
glom <- microbiome::transform(glom, "compositional")
dat <- psmelt(glom)
filt.gen <- dat %>% group_by(Genus) %>% summarize(mean = mean(Abundance)) %>% filter(mean <= 0.01)
dat <- subset(dat, !(OTU %in% filt.gen$Genus))
dat <- dat %>% group_by(Genus) %>% summarize(cs = mean(Abundance)) %>% mutate(cs = cs/sum(cs)) 
dat <- psmelt(glom)
filt.gen <- dat %>% group_by(Genus) %>% summarize(mean = mean(Abundance)) %>% filter(mean <= 0.01)
dat <- subset(dat, !(OTU %in% filt.gen$Genus))
dat <- dat %>% group_by(treatment, Genus) %>% summarize(cs = mean(Abundance)) %>% mutate(cs = cs/sum(cs)) 

library("RColorBrewer")
nb.cols <- length(unique(dat$Genus))
mycolors <- colorRampPalette(brewer.pal(8, "Set3"))(nb.cols)

taxa_plot <- ggplot(dat, aes(x = as.factor(treatment), y = cs, fill = Genus)) +
                        #facet_grid(Site ~ Year) +
                        theme_bw(base_size = 14) +
                        geom_bar(stat="identity") +
                        labs(y = "Relative proportion") +
                        theme(legend.background = element_rect(fill="white"),
                              legend.key = element_rect(fill="transparent"),
                              legend.text = element_text(size = 12),
                              axis.text.x = element_text(color="black"),
                              axis.text.y = element_text(color="black"),
                              panel.grid = element_blank()) +
                        scale_y_continuous(labels = scales::percent) +
                        scale_fill_manual(values = mycolors) +
                        labs(y = "Relative abundance", x="") 
ggsave(taxa_plot, filename = "figures/taxa_ITS.pdf", dpi = 600,  width = 8, height = 5, units = "in")
taxa_plot
```

## 18S
```{r, warning=FALSE, message=FALSE, echo=FALSE}
diagdds = phyloseq_to_deseq2(ps.18s, ~ 1)
ts = counts(diagdds)
geoMeans = apply(ts, 1, function(row) if (all(row == 0)) 0 else exp(mean(log(row[row != 0]))))
diagdds = estimateSizeFactors(diagdds, geoMeans=geoMeans)
diagdds = estimateDispersions(diagdds)
diagvst = getVarianceStabilizedData(diagdds)
diagdds.c <- removeBatchEffect(diagvst)
diagdds.c[diagdds.c<0] <- 0
ps.18sn <- ps.18s
otu_table(ps.18sn) <- otu_table(diagdds.c, taxa_are_rows = TRUE)

glom <- microbiome::aggregate_taxa(ps.18sn, "Genus")
glom <- microbiome::transform(glom, "compositional")
dat <- psmelt(glom)
filt.gen <- dat %>% group_by(Genus) %>% summarize(mean = mean(Abundance)) %>% filter(mean <= 0.01)
dat <- subset(dat, !(OTU %in% filt.gen$Genus))
dat <- dat %>% group_by(Genus) %>% summarize(cs = mean(Abundance)) %>% mutate(cs = cs/sum(cs)) 
dat <- psmelt(glom)
filt.gen <- dat %>% group_by(Genus) %>% summarize(mean = mean(Abundance)) %>% filter(mean <= 0.01)
dat <- subset(dat, !(OTU %in% filt.gen$Genus))
dat <- dat %>% group_by(treatment, Genus) %>% summarize(cs = mean(Abundance)) %>% mutate(cs = cs/sum(cs)) 

library("RColorBrewer")
nb.cols <- length(unique(dat$Genus))
mycolors <- colorRampPalette(brewer.pal(8, "Set3"))(nb.cols)

taxa_plot <- ggplot(dat, aes(x = as.factor(treatment), y = cs, fill = Genus)) +
                        #facet_grid(Site ~ Year) +
                        theme_bw(base_size = 14) +
                        geom_bar(stat="identity") +
                        labs(y = "Relative proportion") +
                        theme(legend.background = element_rect(fill="white"),
                              legend.key = element_rect(fill="transparent"),
                              legend.text = element_text(size = 12),
                              axis.text.x = element_text(color="black"),
                              axis.text.y = element_text(color="black"),
                              panel.grid = element_blank()) +
                        scale_y_continuous(labels = scales::percent) +
                        scale_fill_manual(values = mycolors) +
                        labs(y = "Relative abundance", x="") 
ggsave(taxa_plot, filename = "figures/taxa_18s.pdf", dpi = 600,  width = 8, height = 5, units = "in")
taxa_plot
```

# Differential taxa {.tabset}
## ITS - herbivory
```{r, warning=FALSE, message=FALSE, echo=FALSE}
cal.diff.taxa <- function(object, direction){
  diagdds <- phyloseq_to_deseq2(object, ~ 1)
  ts <- counts(diagdds)
  geoMeans = apply(ts, 1, function(row) if (all(row == 0)) 0 else exp(mean(log(row[row != 0]))))
  diagdds = estimateSizeFactors(diagdds, geoMeans=geoMeans)
  diagdds = estimateDispersions(diagdds)
  diagdds$group <- factor(paste0(diagdds$treatment))
  design(diagdds) <- ~ group
  dds <-DESeq(diagdds, betaPrior=FALSE, parallel = T)
  c1 <- results(dds, contrast=c("group", "herb", "ctr"), parallel = T)
  c1 <- as.data.frame(c1)
  c1 <- setDT(c1, keep.rownames = TRUE)[]
  c1 <- c1[,c("rn", "log2FoldChange", "padj")]
  x <- if(direction == "up"){c1 %>% dplyr::filter(log2FoldChange > 0 & padj < 0.05)} else if(direction == "down"){c1 %>% dplyr::filter(log2FoldChange < 0 & padj < 0.05)}
  tax.table <- as.data.frame(tax_table(object))
  tx <- tax.table[which(row.names(tax.table) %in% x$rn),]
  return(tx)
}

cal.diff.taxa(ps.its, "up") %>% create_dt
```

## ITS - control
```{r, warning=FALSE, message=FALSE, echo=FALSE}
cal.diff.taxa(ps.its, "down") %>% create_dt
```


## 18S - herbivory
```{r, warning=FALSE, message=FALSE, echo=FALSE}
cal.diff.taxa(ps.18s, "up") %>% create_dt
```


## 18S - control
```{r, warning=FALSE, message=FALSE, echo=FALSE}
cal.diff.taxa(ps.18s, "down") %>% create_dt
```

# Volcano plot {.tabset}
## 18S
```{r, warning=FALSE, message=FALSE, echo=FALSE}
cal.diff.taxa2 <- function(object){
  diagdds <- phyloseq_to_deseq2(object, ~ 1)
  ts <- counts(diagdds)
  geoMeans = apply(ts, 1, function(row) if (all(row == 0)) 0 else exp(mean(log(row[row != 0]))))
  diagdds = estimateSizeFactors(diagdds, geoMeans=geoMeans)
  diagdds = estimateDispersions(diagdds)
  diagdds$group <- factor(paste0(diagdds$treatment))
  design(diagdds) <- ~ group
  dds <-DESeq(diagdds, betaPrior=FALSE, parallel = T)
  c1 <- results(dds, contrast=c("group", "herb", "ctr"), parallel = T)
  c1 <- as.data.frame(c1)
  c1 <- setDT(c1, keep.rownames = TRUE)[]
  c1 <- c1[,c("rn", "log2FoldChange", "padj")]
  # x <- if(direction == "up"){c1 %>% dplyr::filter(log2FoldChange > 0 & padj < 0.05)} else if(direction == "down"){c1 %>% dplyr::filter(log2FoldChange < 0 & padj < 0.05)}
  tax.table <- as.data.frame(tax_table(object))
  tax.table <- setDT(tax.table, keep.rownames = TRUE)[]
  tx <- merge(c1, tax.table, by = "rn")
  return(tx)
}

df.diff <- cal.diff.taxa2(ps.18s)

df.diff$diffexpressed <- "no changes"
df.diff$diffexpressed[df.diff$log2FoldChange > 0.1 & df.diff$padj < 0.05] <- "herbivory"
df.diff$diffexpressed[df.diff$log2FoldChange < -0.1 & df.diff$padj < 0.05] <- "control"

plot <- ggplot(data=df.diff) +
        theme_bw(base_size = 12) +
        geom_point(aes(x = log2FoldChange, y = -log10(padj), colour = diffexpressed), size=3) +
        geom_text_repel(aes(x = log2FoldChange, y = -log10(padj), label = ifelse(diffexpressed != "no changes", Species,"")), max.overlaps = 20) +
        scale_color_manual(values=c("blue", "red", "darkgrey")) +
        geom_vline(xintercept=0, col="darkgrey", linetype = "longdash", size=0.5) +
        geom_hline(yintercept=-log10(0.05), col="darkgrey", linetype = "longdash", size=0.5) +
     scale_x_continuous(expand = c(0, 0), limits = c(-30, 30),breaks=seq(-30,30,5))+
   scale_y_continuous(expand = c(0, 0), limits = c(-4, 16),breaks=seq(-4,16,4))+
        theme(legend.justification=c(0.01,0.99), legend.position=c(0.01,0.99),
              panel.grid = element_blank(),
              legend.title=element_blank(), 
              legend.background = element_rect(color = NA),
              legend.key = element_rect(color = NA),
              panel.border = element_rect(colour = "black", fill=NA, size=1)) +
        xlab(expression(paste(Log[2], " Fold Changes"))) +
        ylab(expression(paste(-Log[10], " P"))) +
        scale_color_manual(name = "Legend", values=c("#4daf4a", "#e41a1c", "darkgrey"), labels = c("Control", "Herbivory", "No changes"), breaks = c("control", "herbivory", "no changes"))
ggsave(plot, filename = "figures/volcano_18s.pdf", dpi = 600,  width = 8, height = 5, units = "in")
plot

```

## ITS
```{r, warning=FALSE, message=FALSE, echo=FALSE}
df.diff <- cal.diff.taxa2(ps.its)

df.diff$diffexpressed <- "no changes"
df.diff$diffexpressed[df.diff$log2FoldChange > 0.1 & df.diff$padj < 0.05] <- "herbivory"
df.diff$diffexpressed[df.diff$log2FoldChange < -0.1 & df.diff$padj < 0.05] <- "control"

plot <- ggplot(data=df.diff) +
        theme_bw(base_size = 12) +
        geom_point(aes(x = log2FoldChange, y = -log10(padj), colour = diffexpressed), size=2) +
        geom_text_repel(aes(x = log2FoldChange, y = -log10(padj), label = ifelse(diffexpressed != "no changes", Species,"")), max.overlaps = 20) +
        scale_color_manual(values=c("blue", "red", "black")) +
        geom_vline(xintercept=0, col="darkgrey", linetype = "longdash", size=0.5) +
        geom_hline(yintercept=-log10(0.05), col="darkgrey", linetype = "longdash", size=0.5) +
     scale_x_continuous(expand = c(0, 0), limits = c(-16, 16),breaks=seq(-16,16,4))+
   scale_y_continuous(expand = c(0, 0), limits = c(-4, 16),breaks=seq(-4,16,4))+
        theme(legend.justification=c(0.01,0.99), legend.position=c(0.01,0.99),
              panel.grid = element_blank(),
              legend.title=element_blank(), 
              legend.background = element_rect(color = NA),
              legend.key = element_rect(color = NA),
              panel.border = element_rect(colour = "black", fill=NA, size=1)) +
        xlab(expression(paste(Log[2], " Fold Changes"))) +
        ylab(expression(paste(-Log[10], " P"))) +
        scale_color_manual(name = "Legend", values=c("#4daf4a", "#e41a1c", "darkgrey"), labels = c("Control", "Herbivory", "No changes"), breaks = c("control", "herbivory", "no changes"))
ggsave(plot, filename = "figures/volcano_its.pdf", dpi = 600,  width = 8, height = 5, units = "in")
plot

```






